Finiteness in Cubical Type Theory

### 2020-07-08 sections/topos.tex
fixed agda snippet
Update topos.tex

  We can prove this goal by showing that \(\AgdaFunction{Tuple}\;n\;U\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \agdatop, which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\subsection{Closure on Cardinal Finiteness}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\AgdaDatatype{\ensuremath{\mathcal{B}}}(A)\), closure proofs on \(\AgdaDatatype{\ensuremath{\mathcal{B}}}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) and \(\AgdaDatatype{\ensuremath{\mathcal{B}}}\)).
The cases for \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.

Non-dependent operators like \(\times\), \(\uplus\), and \(\rightarrow\) are
also relatively straightforward: since \(\lVert {\wc} \rVert\) forms a monad, we
can apply \(n\)-ary functions to values inside it, combining them together.
\begin{agdalisting}
  The fact that \(\lVert \wc \rVert\) forms a monad means that we can lift
  \(n\)-ary functions like the following:
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{times-clos-sig}
  Into a truncated context:
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{times-clos-impl}
\end{agdalisting}

Unfortunately, for the dependent type formers like \(\Sigma\) and \(\Pi\), the
same trick does not work.
We have closure proofs like:
\begin{equation}
  \frac{
    \AgdaDatatype{\ensuremath{\mathcal{B}}}(A) \; \; \; \Pi(x : A) , \AgdaDatatype{\ensuremath{\mathcal{B}}}(U(x))
  }{
    \AgdaDatatype{\ensuremath{\mathcal{B}}}(\Pi \; A \; U)
  }
\end{equation}
If we apply the monadic truncation trick we can derive closure proofs like the
following:
\begin{equation}
  \frac{
    \lVert \AgdaDatatype{\ensuremath{\mathcal{B}}}(A) \rVert \; \; \; \lVert \Pi(x : A) , \AgdaDatatype{\ensuremath{\mathcal{B}}}(U(x)) \rVert
  }{
    \lVert \AgdaDatatype{\ensuremath{\mathcal{B}}}(\Pi \; A \; U) \rVert
  }
\end{equation}
However our \emph{desired} closure proof is the following:
\begin{equation}
  \frac{
    \lVert \AgdaDatatype{\ensuremath{\mathcal{B}}}(A) \rVert \; \; \; \Pi(x : A) , \lVert \AgdaDatatype{\ensuremath{\mathcal{B}}}(U(x)) \rVert
  }{
    \lVert \AgdaDatatype{\ensuremath{\mathcal{B}}}(\Pi \; A \; U) \rVert
  }
\end{equation}
They don't match!

The solution would be to find a function of the following type:
\begin{equation}
  (\Pi(x : A) , \lVert \AgdaDatatype{\ensuremath{\mathcal{B}}}(U(x)) \rVert) \rightarrow
  \lVert \Pi(x : A) , \AgdaDatatype{\ensuremath{\mathcal{B}}}(U(x)) \rVert
\end{equation}
However we might be disheartened at realising that this is a required goal: the
above equation is \emph{extremely} similar to the axiom of choice!
\begin{definition}[Axiom of Choice]
  In HoTT, the axiom of choice is commonly defined as follows \cite[lemma
  3.8.2]{hottbook}.
  For any set \(A\), and a type family \(U\) which is a set at all the points
  of \(A\), the following function exists:
  \begin{equation}
    \left( \Pi(x : A) ,  \lVert U(x) \rVert \right) \rightarrow \lVert \Pi(x : A) , U(x) \rVert
  \end{equation}
\end{definition}
Luckily the axiom of choice \emph{does} hold for cardinally finite types,
allowing us to prove the following:
\begin{lemma}
  \begin{equation}
    \mathcal{C}(A) \rightarrow (\Pi(x : A) , \lVert U(x) \rVert) \rightarrow \lVert \Pi(x : A) , U(x) \rVert
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a cardinally finite type, \(U\) be a type family on \(A\), and
  \(f\) be a dependent function of type \(\Pi(x : A) , \lVert U(x) \rVert\).

  First, since our goal is itself propositionally truncated, we have access to
  values under truncations: put another way, in the context of proving our goal,
  we can rely on the fact that \(A\) is manifestly Bishop finite.
  Using the same technique as we did in lemma~\ref{split-enum-pi}, we can switch
  from working with dependent functions from \(A\) to \(n\)-tuples, where \(n\)
  is the cardinality of \(A\).
  This changes our goal to the following:
  \begin{equation}
    \mathbf{Tuple}(n, \lVert \wc \rVert \circ U) \rightarrow \lVert \mathbf{Tuple}(n, U) \rVert
  \end{equation}
  Since \(\lVert \wc \rVert\) is closed under finite products, this function
  exists (in fact, using the fact that \(\lVert \wc \rVert\) forms a monad, we
  can recognise this function as \verb+sequenceA+ from the \verb+Traversable+
  class in Haskell).
\end{proof}

>>>>

  We can prove this goal by showing that \(\AgdaFunction{Tuple}\;n\;U\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \agdatop, which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\subsection{Closure on Cardinal Finiteness}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\AgdaDatatype{\ensuremath{\mathcal{B}}}\;A\), closure proofs on \(\AgdaDatatype{\ensuremath{\mathcal{B}}}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\AgdaDatatype{\ensuremath{\mathcal{E}!}}\) and \(\AgdaDatatype{\ensuremath{\mathcal{B}}}\)).
The cases for \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.

Non-dependent operators like \(\times\), \(\uplus\), and \(\rightarrow\) are
also relatively straightforward: since \(\AgdaDatatype{\ensuremath{\lVert\_\rVert}}\) forms a monad, we
can apply \(n\)-ary functions to values inside it, combining them together.
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{times-clos-sig}
\end{agdalisting}
Into a truncated context:
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{times-clos-impl}
\end{agdalisting}

Unfortunately, for the dependent type formers like \(\Sigma\) and \(\Pi\), the
same trick does not work.
We have closure proofs like:
\begin{equation}
  \frac{
    \AgdaDatatype{\ensuremath{\mathcal{B}}}\;A \; \; \; ((x : A) \rightarrow \AgdaDatatype{\ensuremath{\mathcal{B}}}\;(U\;x))
  }{
    \AgdaDatatype{\ensuremath{\mathcal{B}}}\;((x : A) \rightarrow U\;x)
  }
\end{equation}
If we apply the monadic truncation trick we can derive closure proofs like the
following:
\begin{equation}
  \frac{
    \AgdaDatatype{\ensuremath{\lVert}}\; \AgdaDatatype{\ensuremath{\mathcal{B}}}\;A \;\AgdaDatatype{\ensuremath{\rVert}} \; \; \; \AgdaDatatype{\ensuremath{\lVert}}\; ((x : A) \rightarrow \AgdaDatatype{\ensuremath{\mathcal{B}}}\;(U\;x)) \;\AgdaDatatype{\ensuremath{\rVert}}
  }{
    \AgdaDatatype{\ensuremath{\lVert}}\; \AgdaDatatype{\ensuremath{\mathcal{B}}}\;((x : A) \rightarrow U\;x) \;\AgdaDatatype{\ensuremath{\rVert}}
  }
\end{equation}
However our \emph{desired} closure proof is the following:
\begin{equation}
  \frac{
    \AgdaDatatype{\ensuremath{\lVert}}\; \AgdaDatatype{\ensuremath{\mathcal{B}}}\;A \;\AgdaDatatype{\ensuremath{\rVert}} \; \; \; ((x : A) \rightarrow \AgdaDatatype{\ensuremath{\lVert}}\; \AgdaDatatype{\ensuremath{\mathcal{B}}}\;(U\;x) \;\AgdaDatatype{\ensuremath{\rVert}})
  }{
    \AgdaDatatype{\ensuremath{\lVert}}\; \AgdaDatatype{\ensuremath{\mathcal{B}}}\;((x : A) \rightarrow U\;x) \;\AgdaDatatype{\ensuremath{\rVert}}
  }
\end{equation}
They don't match!

The solution would be to find a function of the following type:
\begin{equation}
  ((x : A) \rightarrow \AgdaDatatype{\ensuremath{\lVert}}\; \AgdaDatatype{\ensuremath{\mathcal{B}}}\;(U\;x) \;\AgdaDatatype{\ensuremath{\rVert}}) \rightarrow
  \AgdaDatatype{\ensuremath{\lVert}}\; (x : A) \rightarrow \AgdaDatatype{\ensuremath{\mathcal{B}}}\;(U\;x) \;\AgdaDatatype{\ensuremath{\rVert}}
\end{equation}
However we might be disheartened at realising that this is a required goal: the
above equation is \emph{extremely} similar to the axiom of choice!
\begin{definition}[Axiom of Choice]
  In HoTT, the axiom of choice is commonly defined as follows \cite[lemma
  3.8.2]{hottbook}.
  For any set \(A\), and a type family \(U\) which is a set at all the points
  of \(A\), the following function exists:
  \begin{equation}
    \left( (x : A) \rightarrow  \AgdaDatatype{\ensuremath{\lVert}}\; U(x) \;\AgdaDatatype{\ensuremath{\rVert}} \right) \rightarrow \AgdaDatatype{\ensuremath{\lVert}}\; (x : A) \rightarrow U(x) \;\AgdaDatatype{\ensuremath{\rVert}}
  \end{equation}
\end{definition}
Luckily the axiom of choice \emph{does} hold for cardinally finite types,
allowing us to prove the following:
\begin{lemma}
  \begin{equation}
    \agdacal{C}\;A \rightarrow ((x : A) \rightarrow \AgdaDatatype{\ensuremath{\lVert}}\; U(x) \;\AgdaDatatype{\ensuremath{\rVert}}) \rightarrow \AgdaDatatype{\ensuremath{\lVert}}\; (x : A) \rightarrow U(x) \;\AgdaDatatype{\ensuremath{\rVert}}
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a cardinally finite type, \(U\) be a type family on \(A\), and
  \(f\) be a dependent function of type \(\Pi(x : A) , \AgdaDatatype{\ensuremath{\lVert}}\; U(x) \;\AgdaDatatype{\ensuremath{\rVert}}\).

  First, since our goal is itself propositionally truncated, we have access to
  values under truncations: put another way, in the context of proving our goal,
  we can rely on the fact that \(A\) is manifestly Bishop finite.
  Using the same technique as we did in lemma~\ref{split-enum-pi}, we can switch
  from working with dependent functions from \(A\) to \(n\)-tuples, where \(n\)
  is the cardinality of \(A\).
  This changes our goal to the following:
  \begin{equation}
    \mathbf{Tuple}(n, \AgdaDatatype{\ensuremath{\lVert}}\; \wc \;\AgdaDatatype{\ensuremath{\rVert}} \circ U) \rightarrow \AgdaDatatype{\ensuremath{\lVert}}\; \mathbf{Tuple}(n, U) \;\AgdaDatatype{\ensuremath{\rVert}}
  \end{equation}
  Since \(\AgdaDatatype{\ensuremath{\lVert\_ \rVert}}\) is closed under finite products, this function
  exists (in fact, using the fact that \(\AgdaDatatype{\ensuremath{\lVert\_ \rVert}}\) forms a monad, we
  can recognise this function as \verb+sequenceA+ from the \verb+Traversable+
  class in Haskell).
\end{proof}
